---
title: "M7: Forecasting in R: Simple Averaging Models & ARIMA"
author: "Luana Lima"
output: pdf_document
editor_options: 
  chunk_output_type: console
---

## Setting R code chunk options

First R code chunk is used for setting the options for all R code chunks. The choice echo=TRUE means both code and output will appear on report, include = FALSE neither code nor output is printed.

```{r setup, include=FALSE} 
knitr::opts_chunk$set(echo = TRUE,tidy.opts=list(width.cutoff=80), tidy=FALSE) 
```

## Loading packages and initializing

Second R code chunk is for loading packages. By setting message = FALSE, the code will appear but not the output. 

```{r package, message=FALSE}
library(lubridate)
library(ggplot2)
library(forecast)  
library(Kendall)
library(tseries)
library(outliers)
library(tidyverse)

#install.packages("smooth")
library(smooth)

```

## Importing data

For this module we will continue to work with the electricity retail price in US dataset from the U.S. Energy Information Administration. You may download  [here][https://www.eia.gov/electricity/data/browser/#/topic/7?agg=2,0,1&geo=g&freq=M%2013:41:41%20GMT-0500%20(EST)]. But this week we will work with the all.sectors column instead of residential price.

```{r}

#Importing time series data from text file#
electricity_price <- read.csv(file="./Data/Average_retail_price_of_electricity_United_States_monthly.csv",header=TRUE,skip=4)

#Inspect data
head(electricity_price)
nvar <- ncol(electricity_price) - 1
nobs <- nrow(electricity_price)

#Preparing the data - create date object and rename columns
electricity_price_processed <-
  electricity_price %>%
  mutate( Month = my(Month) ) %>% 
  rename( All.sectors = all.sectors.cents.per.kilowatthour ) %>% 
  rename( Residential = residential.cents.per.kilowatthour ) %>% 
  rename( Commercial = commercial.cents.per.kilowatthour ) %>% 
  rename( Industrial = industrial.cents.per.kilowatthour ) %>% 
  arrange( Month )

head(electricity_price_processed)
summary(electricity_price_processed)

#No NAs so we don't need to worry about missing values

```

## Transforming data into time series object

Many of the functions we will use require a time series object. You can transform your data in a time series using the function *ts()*. 

```{r}
ts_electricity_price <- ts(electricity_price_processed[,2:(nvar+1)],
                           start=c(year(electricity_price_processed$Month[1]),month(electricity_price_processed$Month[1])),
                           frequency=12) 
#note that we are only transforming columns with electricity price, not the date columns  
head(ts_electricity_price,15)
tail(ts_electricity_price,15)

```

## Initial Plots

```{r}

TS_Plot <- 
  ggplot(electricity_price_processed, aes(x=Month, y=All.sectors)) +
      geom_line()
plot(TS_Plot)

#ACF and PACF plots
par(mfrow=c(1,2))
ACF_Plot <- Acf(electricity_price_processed$All.sectors, lag = 40, plot = TRUE)
PACF_Plot <- Pacf(electricity_price_processed$All.sectors, lag = 40)
par(mfrow=c(1,1))
```


## Decomposing the time series 

The plots from the previous section show the data has a seasonal component. Since we are working with non-seasonal ARIMA, we need to decompose the series and eliminate the seasonality.

```{r}
#Using R decompose function
decompose_allsectors_price <- decompose(ts_electricity_price[,"All.sectors"],"additive")
plot(decompose_allsectors_price)

#The ACF plot show a slow decay which is a sign of non-stationarity.

#Creating non-seasonal residential price time series because some models can't handle seasonality
deseasonal_allsectors_price <- seasadj(decompose_allsectors_price)  


```


## Forecasting with Naive methods

We will go over three naive methods.
(a) Forecast using the arithmetic mean with function meanf() from package `forecast`.
(b) Forecast using the naive method with function naive() from package `forecast`.
(c) Forecast using the seasonal naive method with function snaive() from package `forecast`. You need to specify frequency when defining the ts object to be able to use this function.

These three function take a few main arguments:

**y** ts object containing data needed to be forecasted
**h** forecasting horizon, i.e., how many steps ahead do you want to forecast
**level** Confidence level for prediction intervals



```{r}
#Model 1: Arithmetic mean on original data
Model1 <-meanf(ts_electricity_price[,"All.sectors"],
               h=12)
summary(Model1)
plot(Model1)
# use the original time series data;
#summary gets errors and forecasts that is the same as the mean of the series - we have trend and seasonality and the mean is a bad approach.

#Model 2: Arithmetic mean on deseas data
Model2 <-meanf(deseasonal_allsectors_price,
               h=12)
summary(Model2)
plot(Model2)
# average mean doesn't work either because there is a trend;

#Model 3: Seasonal naive on original data
Model3 <-snaive(ts_electricity_price[,"All.sectors"],
               h=12)
summary(Model3)
plot(Model3)
# is better than the overall mean! if you have a trend; it is better to repeat the previous information because there is a trend
#Model 4: Naive on deseas data
Model4 <-naive(deseasonal_allsectors_price, h=12)
summary(Model4)
plot(Model4)
# considering last year but without seasonality
# training set: just the forecast errors for training period
#not doing any holdout on this yet; that is the performance criteria - use the entire model
```



## Forecasting with Simple Averaging Models

The simple averaging techniques we will explore are the moving average and the exponential smoothing. The simple moving average can be obtained with function sma() from package `smooth`. The main arguments function sma() will take are:

**y** ts object containing data needed to be forecasted
**order** order of simple moving average. If you don't specify, R will find the best order for you
**h** forecasting horizon, i.e., how many steps ahead do you want to forecast
**holdout** If TRUE, the last **h** observations are taken from the end of the data. One reason to holdout some data is to check model performance. We will talk about that on M9.
**silent** if FALSE, no output will be generated
**level** Confidence level for prediction intervals


The simple exponential smoothing can be obtained with the ses() function also from package `forecast`. The main arguments function ses() will take are:

**y** ts object containing data needed to be forecasted
**alpha** the smoothing constant. If you don't specify, R will find the best alpha for you. 
**h** forecasting horizon, i.e., how many steps ahead do you want to forecast
**holdout** If TRUE, the last **h** observations are taken from the end of the data. One reason to holdout some data is to check model performance. We will talk about that on M9.
**silent** if FALSE, no output will be generated
**level** Confidence level for prediction intervals


We will try these models on the original dataset and the deaseason electricity price.

```{r}
# Model 5: Simple moving average on deseasonal data
Model5 <- smooth::sma(deseasonal_allsectors_price, h=12)

# for sma: hold out works! holdout = TRUE for Model 5
#have to do the model by hand because sma is from package smooth
autoplot(deseasonal_allsectors_price,series = "Deaseasonal") +
  autolayer(Model5$forecast, series="SMA")


# Model 6: Simple moving average on original data
Model6 <- sma(ts_electricity_price[,"All.sectors"], h=12)
#summary(Model6)
#have to do the model by hand because sma is from package smooth
autoplot(ts_electricity_price[,"All.sectors"], series = "Original") +
  autolayer(Model6$forecast, series="SMA")

# Model 7:  Simple exponential smoothing on deseasonal data
Model7 <- ses(deseasonal_allsectors_price, h=12)
plot(Model7)
# could detrend ... but then would have to add back in
# Model 8:  Simple exponential smoothing on original data
Model8 <- ses(ts_electricity_price[,"All.sectors"],h=12)
plot(Model8)

# all these models look the same; not observing anything; assuming errors are 0
# these models can't produce forecasts that are good enough
autoplot(ts_electricity_price[,"All.sectors"],
         series = "Original") +
  autolayer(Model8$fitted, series = "ES-Model8")
# exponential smoothing: has a delay; same as the last obs; so it will become a straight line
# always trying to fix the errors to be better; this is going to be relatively good for it to be just one step ahead
```


## Forecasting with ARIMA

When forecasting with ARIMA models, first you will need to fit the model with the auto.arima() function we studied on M6 and M7. Then you will need to use the forecast() function from package `forecast`. forecast() is a generic function for forecasting from time series or time series models. It takes two main arguments:

**object** 	a time series or time series model for which forecasts are required;
**h** forecasting horizon, i.e., how many steps ahead do you want to forecast.

```{r}

# Model 9:  SARIMA on original data
Model9 <- auto.arima(ts_electricity_price[,"All.sectors"])
# just an arima fit; not holdout, just estimate the order and parameters for the arima for auto.arima
Model9_for <- forecast(Model9, h=12)
plot(Model9_for)

autoplot(ts_electricity_price[,"All.sectors"],
         series = "Original") +
  autolayer(Model9$fitted, series = "SARIMA") +
  autolayer(Model9_for$mean, series = "SARIMA_for")
# works fine with autoplot because it knows the time
summary(Model9_for)
#autoplot is ts objects; so date works fine; if doing with ggplot or plot - BE CAREFUL!

# Model 10:  ARIMA on deseasonal data
Model10 <- auto.arima(deseasonal_allsectors_price, seasonal = FALSE)
Model10_for <- forecast(Model10, h=12)

autoplot(deseasonal_allsectors_price,
         seasonal = FALSE,
         series = "Deseasoned") +
  autolayer(Model10$fitted, series = "ARIMA") +
  autolayer(Model10_for$mean, series = "ARIMA_for")
# arima will model the trend!
# don't compare arima and sarima errors ! different data sets!
```


