---
title: "ENV 790.30 - Time Series Analysis for Energy Data | Spring 2024"
subtitle: "Assignment 3 - Due date 02/01/24"
author: "Samantha Pace"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## Directions
  
You should open the .rmd file corresponding to this assignment on RStudio. The file is available on our class repository on Github. 

Once you have the file open on your local machine the first thing you will do is rename the file such that it includes your first and last name (e.g., "LuanaLima_TSA_A02_Sp24.Rmd"). Then change "Student Name" on line 4 with your name. 

Then you will start working through the assignment by  **creating code and output** that answer each question. Be sure to use this assignment document. Your report should contain the answer to each question and any plots/tables you obtained (when applicable).

Please keep this R code chunk options for the report. It is easier for us to grade when we can see code and output together. And the tidy.opts will make sure that line breaks on your code chunks are automatically added for better visualization. 

```{r setup, include=FALSE} 
knitr::opts_chunk$set(echo = TRUE,tidy.opts=list(width.cutoff=80), tidy=FALSE) 
```

When you have completed the assignment, **Knit** the text and code into a single PDF file. Submit this pdf using Sakai. 

## Questions 

Consider the same data you used for A2 from the spreadsheet "Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx". The data comes from the US Energy Information and Administration and corresponds to the December 2022 **Monthly** Energy Review. Once again you will work only with the following columns: Total Biomass Energy Production, Total Renewable Energy Production, Hydroelectric Power Consumption. Create a data frame structure with these three time series only. 

R packages needed for this assignment:"forecast","tseries", and "Kendall". Install these packages, if you haven't done yet. Do not forget to load them before running your script, since they are NOT default packages.\\

```{r}
#Load/install required package here

#install.packages("forecast")
#install.packages("tseries")
#install.packages("Kendall")
#install.packages("cowplot")
#install.packages("ggplot2")
#install.packages("dplyr")

library(forecast)
library(tseries)
library(Kendall)
library(readxl)
library(cowplot)
library(ggplot2)
library(dplyr)

#Importing data using read.xlsx
energy_data <- read_excel(
  path="./Data/Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx", 
                         skip = 12,
                         sheet = "Monthly Data", col_names = FALSE)

# Getting column names from row 11
read_col_names <-
  read_excel(path="./Data/Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx",
             skip = 10,n_max = 1, sheet="Monthly Data",col_names=FALSE)

energy_data <- 
  read_excel(path="./Data/Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx",
             skip = 12, sheet="Monthly Data",col_names=FALSE) 

# inputting correct column names
colnames(energy_data) <- read_col_names
head(energy_data)

# selecting the 3 columns needed
energy_data <- energy_data %>% select(1, 5, 6)
```

##Trend Component

### Q1
For each time series, i.e., Renewable Energy Production and Hydroelectric Consumption create three plots: one with time series, one with the ACF and with the PACF. You may use the some code form A2, but I want all the three plots side by side as in a grid.
(Hint: use function `plot_grid()` from the `cowplot` package)

```{r Question 1}
# transforming data frame in a time series object
REP_ts <- ts(energy_data$`Total Renewable Energy Production`, start = c(1973,1), frequency = 12)
HEC_ts <- ts(energy_data$`Hydroelectric Power Consumption`, start = c(1973,1), frequency = 12)

# creating individual plots for Renewable Energy Production
REP_plot1 <- autoplot(REP_ts,
                  ylab= "Energy (Btu)",
                  main = "Renewable Energy Production over Time")

REP_ACF <- acf(REP_ts, 
               col = 'blue',
               lag.max = 40, 
               ylab = "ACF", 
               main = "ACF of Renewable Energy")

REP_ACF_autoplot <- autoplot(REP_ACF)

REP_PACF <- pacf(REP_ts, 
                 col = 'purple',
                 lag.max = 40, 
                 ylab = 'PACF', 
                 main = "PACF of Renewable Energy")

REP_PACF_autoplot <- autoplot(REP_PACF)

# combining REP individual with plot grid
plot_grid(REP_plot1, REP_ACF_autoplot, REP_PACF_autoplot)

# creating individual plots for Hydroelectric Consumption
HEC_plot1 <- autoplot(HEC_ts,
                ylab = "Energy (Btu)",
                main = "Hydroelectric Energy Consumption over Time")

HEC_ACF <- acf(HEC_ts,
               lag.max = 40, 
               ylab = "ACF", 
               main = "ACF of Hydroelectric Consumption")

HEC_ACF_autoplot <- autoplot(HEC_ACF)

HEC_PACF <- pacf(HEC_ts, 
                 lag.max = 40, 
                 ylab = "PACF", 
                 main = "PACF of Hydroelectric Consumption")

HEC_PACF_autoplot <- autoplot(HEC_PACF)

# combining with plotgrid
plot_grid(HEC_plot1, HEC_ACF_autoplot, HEC_PACF_autoplot)

```


### Q2
From the plot in Q1, do the series Total Biomass Energy Production, Total Renewable Energy Production, Hydroelectric Power Consumption appear to have a trend? If yes, what kind of trend?

Answer: Both the Total Renewable Energy Production and Hydroelectric Consumption appear to have trends from their plots. The Renewable Energy data appears to have an increasing trend, while the Hydroelectric Energy appears to have a slight decreasing trend. The Hydroelectric ACF graph clearly shows a regularly spaced seasonality component as well.

### Q3
Use the *lm()* function to fit a linear trend to the three time series. Ask R to print the summary of the regression. Interpret the regression output, i.e., slope and intercept. Save the regression coefficients for further analysis.

```{r Question 3}
# storing vectors
nobs <-nrow(energy_data)
t <- 1:nobs

# fitting linear trend to Renewable Energy Production Data
REP_linear_trend <- lm(energy_data$`Total Renewable Energy Production`~t)
summary(REP_linear_trend)

# saving the regression coefficients for REP
REP_beta0 <- REP_linear_trend$coefficients[1]
REP_beta1 <- REP_linear_trend$coefficients[2]

# fitting linear trend to Hydroelectric Consumption (HEC) Data
HEC_linear_trend <- lm(energy_data$`Hydroelectric Power Consumption`~t)
summary(HEC_linear_trend)

# saving the regression coefficients
HEC_beta0 <- HEC_linear_trend$coefficients[1]
HEC_beta1 <- HEC_linear_trend$coefficients[2]

```

Answer: For Renewable Energy, the coefficient of t is 0.704, and has a positive slope, which means the linear model is increasing. The intercept is 180.989. The p-value is less than 0.05, meaning it is statistically significant. The adjusted R squared value is 0.808, which means that the model accounts for 80.8% of the variation of the data.

For Hydroelectric Energy, the coefficient of t is -0.00985, which is a slightly decreasing because it is negative. The intercept is 82.73. The p value is less than 0.05, so the findings are statistically significant. The Adjusted R-squared value is 0.013, which means this model, specifically the component of time, only accounts for 1.3% of the variability of the data, which is very low.
 
### Q4
Use the regression coefficients from Q3 to detrend the series. Plot the detrended series and compare with the plots from Q1. What happened? Did anything change?

```{r}
# Creating REP detrended series
REP_detrend <- energy_data[,2] - (REP_beta0 + REP_beta1*t)

colnames(REP_detrend)[1]<-"detrend"

REP_detrend_df <- data.frame("Month" = energy_data$Month,
                             "Observed" = energy_data[,2],
                             "Detrend" = REP_detrend)

#plotting observed and detrended REP series
ggplot(REP_detrend_df, aes(x=Month)) +
  geom_line(aes(y=Total.Renewable.Energy.Production), color = 'black') +
  geom_line(aes(y=detrend), color = 'purple') +
  ggtitle("Observed and Detrended Renewable Energy")

# Creating HEC detrended series
HEC_detrend <- energy_data[,3] - (HEC_beta0 + HEC_beta1*t)

colnames(HEC_detrend)[1]<-"detrend"

HEC_detrend_df <- data.frame("Month" = energy_data$Month,
                             "Observed" = energy_data[,3],
                             "Detrend" = HEC_detrend)

#plotting observed and detrended HEC series
ggplot(HEC_detrend_df, aes(x=Month)) +
  geom_line(aes(y=Hydroelectric.Power.Consumption), color = 'black') +
  geom_line(aes(y=detrend), color = 'darkgreen') +
  ggtitle("Observed and Detrended Hydroelectric Energy Consumption")

```

Answer: For the Renewable Energy plot, the data is now centered on zero since the intercept was removed. The positive linear model was also removed, making the slope of the detrended data less steep. The observations for the detrended data now range from aboout -125 Btu to 125 Btu. For the Hydroelectric Consumption comparison of observed and detrended, since there was very little slope, the main noticable change is the removal of the intercept from the detrended series which centers the data around zero. The rest of the variability of the dataset looks to be nearly the same.

### Q5
Plot ACF and PACF for the detrended series and compare with the plots from Q1. You may use plot_grid() again to get them side by side. nut mot mandatory. Did the plots change? How?

```{r}
# transforming data frame in a time series object
REP_detrend_ts <- ts(REP_detrend_df$detrend, start = c(1973,1), frequency = 12)
HEC_detrend_ts <- ts(HEC_detrend_df$detrend, start = c(1973,1), frequency = 12)

# plotting detrended ACF and PACF for Renewable Energy
REP_detrend_ACF <- acf(REP_detrend_ts, 
               lag.max = 40, 
               ylab = "ACF", 
               main = "ACF of DETRENDED Renewable Energy")

REP_ACF_detrend_autoplot <- autoplot(REP_detrend_ACF)

plot_grid(HEC_ACF_autoplot, HEC_ACF_detrend_autoplot)

REP_detrend_PACF <- pacf(REP_detrend_ts,
                 lag.max = 40, 
                 ylab = 'PACF', 
                 main = "PACF of DETRENDED Renewable Energy")

REP_PACF_detrend_autoplot <- autoplot(REP_detrend_PACF)

# creating individual plots for Hydroelectric Consumption
HEC_detrend_ACF <- acf(HEC_detrend_ts,
               lag.max = 40, 
               ylab = "ACF", 
               main = "ACF of DETRENDED Hydroelectric Consumption")

HEC_ACF_detrend_autoplot <- autoplot(HEC_detrend_ACF)

HEC_detrend_PACF <- pacf(HEC_detrend_ts, 
                 lag.max = 40, 
                 ylab = "PACF", 
                 main = "PACF of DETRENDED Hydroelectric Consumption")

HEC_PACF_detrend_autoplot <- autoplot(HEC_detrend_PACF)

```

Answer: The Renewable Energy's detrended compared to the original ACF shows a slightly lower and more irregular dependence on t from one lag to the next. The PACF of the observed and detrended Renewable Energy looks fairly similar, with a similar pattern of future lags having a coefficient still.
For the Hydroelectric Energy Consumption, there looks to be the same pattern of seasonality in the ACF for the original and detrended data.

## Seasonal Component

Set aside the detrended series and consider the original series again from Q1 to answer Q6 to Q8.


### Q6
Just by looking at the time series and the acf plots, do the series seem to have a seasonal trend? No need to run any code to answer your question. Just type in you answer below.

### Q7
Use function *lm()* to fit a seasonal means model (i.e. using the seasonal dummies) the two time series. Ask R to print the summary of the regression. Interpret the regression output. From the results which series have a seasonal trend? Do the results match you answer to Q6?

### Q8
Use the regression coefficients from Q7 to deseason the series. Plot the deseason series and compare with the plots from part Q1. Did anything change?

### Q9
Plot ACF and PACF for the deseason series and compare with the plots from Q1. You may use plot_grid() again to get them side by side. nut mot mandatory. Did the plots change? How?


